Cluster 0 -> med

So antihistamines can cause weight gain. NOW they tell me. :-) Is there any way
to find out which do & which don't? My doctor obviously is asleep at the wheel.
The original poster mentioned fatigue. I had that too, but it was mostly due to
the really bizarre dreams I was having -- I wasn't getting any rest. My doctor
said that was a common reaction. If astemizole doesn't cross the blood-brain
barrier, how does it cause that side effect? Any ideas? --

Cluster 1 -> autos

If it looks like a miniature corvette it would be an opel GT,the headlights are
flipped over by pulling a lever inside.>

Cluster 2 -> graphics (computer)

I am currently using POVRay on Mac and was wondering if anyone in netland knows
of public domain anti-aliasing utilities so that I can skip this step in POV,
very slow on this machine. Any suggestions, opinions about post-trace
anti-aliasing would be greatly appreciated.

Cluster 3 -> graphics (geometry)

I am currently using POVRay on Mac and was wondering if anyone in netland knows
of public domain anti-aliasing utilities so that I can skip this step in POV,
very slow on this machine. Any suggestions, opinions about post-trace
anti-aliasing would be greatly appreciated.

Cluster 4 -> graphics (general: merging of the other two)

Hi there, I am looking for advice on software/hardware package for making,
storing and processing of pictures. The ideal software would allow me to cahnge
size of the picture, edit it ( it means add text below, above...) and the most
important is it would have DOS command interface... Thank you in advance...
emanuel marciniak the bank of new york..

cluster 5 -> autos (politcs) I remember this happening on the I-75 through
Michigan and Ohio several years back. A group of guys in an old beater would
rear end a car, usually out of state or Canadians. You stop and they smack you
with a BB bat. At least they didn't kill you for the sake of a car. I think the
cops put out decoys and this calmed down for a while.

Cluster 6 -> med 0 + graphics 4

Is the ".3ds" file format for Autodesk's 3D Animation Studio available? Thanks,
Gary

Cluster 7 -> hockey

Roberts played in last night game against the Sharks and got a goal (38th) and an assist. This definitely bolsters Calgary's chances in the playoffs. Rahim Hirji